EuPPollNet: A European database of plant-pollinator networks

### 2024-09-05 Manuscript/EuPPollNet_Ms.qmd
running everything with the most updated data

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent as defined by the European Environment Agency [@stanners1995]. As this database is the result of one of the work packages of the European project Safeguard (Safeguarding European wild pollinators; https://doi.org/10.3030/101003476), first, data was directly requested from members of the Safeguard project in May 2022. Second, the request was extended to data owners outside of the project, with data collection concluding in August 2024. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. While Google Scholar lacks reproducibility [@gusenbauer2020], it still remains the most comprehensive search engine to date [@gusenbauer2019]. This approach maximized the potential number of studies that could be incorporated in this database. The search strings used were "*plant-pollinator interactions*" and "*plant-pollinator networks*". To maintain high quality standards that will support robust future ecological research, we only included studies that met the following criteria: 1) studies containing time- and geo-referenced records of plant-pollinator interactions; and 2) studies that quantify interactions by documenting them as the contact between a pollinator and the reproductive structure of a specific plant being sampled (i.e., phyto-centric networks).

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we evaluated the taxonomic coverage of taxonomic groups with species checklists available in Europe (i.e., bees, syrphids and butterflies). While there is not a good understanding of pollinator diversity in other taxonomic groups, it was assumed that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`). Therefore, the total number of of flower visiting species from other taxonomic groups (i.e., non-bee, non syrphid and non-butterfly flower visitors) at European level was extrapolated by assuming that their coverage is equal to the mean coverage of bees, syrphids, and butterflies. Consequently, we provide an estimate for the total number of pollinators across the European continent.

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness for each network, and examined how these network metrics change across latitude of studies and bioclimatic regions in Europe. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta-regression implemented with the package **betareg** [@cribari2010]. We used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we quantified the association between connectance and nestedness with the number of species per network using a Kendall rank correlation coefficient to compare the strength of associations between network structures and species richness across both network metrics.

@@

EuPPollNet offers the largest set of plant-pollinator studies and networks compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species).

This likely reflects the fact that most plant and pollinator species are rare and geographically restricted. For example given that most of the plant-pollinator networks from the database are sampled on intensive grasslands, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Indeed, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

>>>>

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent as defined by the European Environment Agency [@stanners1995]. As this database is the result of one of the work packages of the European project Safeguard (Safeguarding European wild pollinators; https://doi.org/10.3030/101003476), first, data was directly requested from members of the Safeguard project in May 2022. Second, the request was extended to data owners outside of the project, with data collection concluding in August 2024. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. While Google Scholar lacks reproducibility [@gusenbauer2020], it still remains the most comprehensive search engine to date [@gusenbauer2019]. This approach maximized the potential number of studies that could be incorporated in this database. The search strings used were "*plant-pollinator interactions*" and "*plant-pollinator networks*". To maintain high quality standards that will support robust future ecological research, we only included studies that met the following criteria: 1) studies containing time- and geo-referenced records of plant-pollinator interactions; and 2) studies that quantify interactions by documenting the contact between a floral visitor - referred to as a 'pollinator' throughout the manuscript, even though the pollination efficiency is not evaluated - and the reproductive structure of a specific sampled plant (i.e., phyto-centric networks).

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we only evaluated the taxonomic coverage of groups with species checklists available in Europe (i.e., bees, syrphids and butterflies). While there is not a good understanding of pollinator diversity in other taxonomic groups (e.g., beetles), it was assumed that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`). Therefore, the total number of of flower visiting species from other taxonomic groups (i.e., non-bee, non syrphid and non-butterfly flower visitors) at European level was extrapolated by assuming that their coverage is equal to the mean coverage of bees, syrphids, and butterflies. Consequently, we provide an estimate for the total number of pollinators across the European continent.

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness for each network in binary form (i.e., presence-absence) and examined how these network metrics change across different latitudes and bioclimatic regions in Europe. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using beta regression. This approach was chosen because connectance displays a non-normal distribution with continuous values bounded between 0 and 1. The model was implemented using the betareg package [@cribari2010]. We used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, to quantify how connectance and nestedness change with network size, we determined their association with the number of species per network using the Kendall rank correlation coefficient.

@@

EuPPollNet offers the largest set of plant-pollinator studies and networks compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). However, although our study likely has a higher pollinator coverage, as we did not exclude non-pollinating species (e.g., predators, herbivores, parasites) from the checklists, this likely reflects that most plant and pollinator species are rare and geographically restricted. For example given that most of the plant-pollinator networks from the database are sampled on intensive grasslands, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Indeed, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require numerous sampling events within and across habitats, particularly for plant-pollinator interactions.
